Mathematical Aspects of the Capital Market

### 2014-02-07 thesis/thesis.tex
A lot of things...

\documentclass{book}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{amssymb}
\usepackage{enumerate}
\usepackage[bookmarks=true]{hyperref}
\usepackage{bookmark}

@@

\chapter{Introduction}
$\mathbb{E}(x)$

\section{Stylized facts}

\section{GARCH models}
The autocorrelation function of $\epsilon_t^2$ is given by
(c.f. \cite{Bollerslev87})
$$
\rho_n = \sum_{i=1}^{p \vee q} (\alpha_i + \beta_i) \rho_{n-i}
\;n > p
$$
where $\alpha_i$ with $i > q$ and $\beta_i$ with $i > p$ are taken as
zeros. $p \vee q$ denotes the maximum of p and q. From these
equations, it is clear that the partial autocorrelation function cuts
off at $max(p, q)$.

\chapter{Case Study of Stocks}
\section{Nordea Bank}
\begin{figure}[ht]
  \centering
  \includegraphics[scale=0.5, clip=true, trim=103 236 118
  200]{../pics/nordea_price_20120116-20120420.pdf}
  \caption{Nordea Bank paid prices 2012/01/16-2012/04/20}
  \label{fig:Nordea}
\end{figure}
Figure \ref{fig:Nordea} shows the paid prices of Nordea Bank from
2012/01/16 to 2012/04/20.

\subsection{15min Returns}
\subsubsection{GARCH model}
The autocorrelation function (ACF) of $\epsilon_t^2$ are shown in figure
\ref{fig:nordea_15min_acf}. At the absence of GARCH effects, the ACF
will have an asymptotic Gaussian distribution with mean 0 and variance
1/T \cite{Bollerslev86, Bollerslev87}. The first 5 autocorrelations
have considerable sizes and do not fall off as in the Gaussian
case. This observation suggests a GARCH model. Moreover, from
\ref{fig:nordea_15min_vlt_acf}, one sees apparent autocorrelations
between volatilities of consecutive time intervals. This again points
to a GARCH model.
\begin{figure}[htb!]
  \centering
  \subfigure[ACF of squared returns]{
    \includegraphics[scale=0.4, clip=true, trim=95 236 118
    200]{../pics/nordea_15min_acf.pdf}
    \label{fig:nordea_15min_acf}
  }
  \subfigure[ACF of volatilities]{
    \includegraphics[scale=0.4, clip=true, trim=95 236 118
    200]{../pics/nordea_15min_vlt_acf.pdf}
    \label{fig:nordea_15min_vlt_acf}
  }
  \caption{Nordea Bank 15min autocorrelations}
\end{figure}

To obtain the functional form of the conditional distribution of
$\epsilon_t$ \footnote{We use the realized volatility as an estimate of the
conditional volatility. See \cite{Andersen03}.}, we look at the
qq-plot of $\epsilon_t/\sigma_t$ \ref{fig:nordea_15min_epOversig_qq},
which is consistent with a standard Gaussian distribution except for a
few outliers at the two ends. Hence the fat tails of the probability
density function can be entirely accounted for by GARCH effects,
implying a Gaussian conditional distribution is appropriate for $\epsilon_t$.
\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.6, clip=true, trim=100 236 118
    200]{../pics/nordea_15min_epOversig_qq.pdf}
  \caption{QQ-plot of $\epsilon_t / \sigma_t$. $\epsilon_t$ are
    derived from Nordea Bank 15min returns while $\sigma_t$ are realized
    volatilities calculated using 30s returns within each 15min
    interval.}
  \label{fig:nordea_15min_epOversig_qq}
\end{figure}

To determine the orders p and q of the GARCH model, we again check
the partial autocorrelation function of $\epsilon_t^2$. As just
mentioned, the first 5 partial autocorrelations are significant. Thus
we can tentatively estimate a GARCH(1, 5) model. Then by applying the
Lagrange multiplier test and accordingly removing insignificant
parameters, we obtain a GARCH(1, 3) model
\ref{tab:nordea_15min_garch}:
\begin{table}[htb!]
  \centering
  \begin{tabular}{c|c|c|c}
    parameter & value & std. error & t statistic \\
    \hline
     $\alpha_0$ &   1.05668e-06 &   3.56871e-07 &    2.96096\\
     \hline
     $\beta_1$  &     0.687212  &   0.0299314   &    22.9596\\
     \hline
     $\alpha_1$ &      0.128614 &   0.0268144   &    4.79643\\
     \hline
     $\alpha_3$ &    0.0568022  &   0.0301503   &    1.88397
  \end{tabular}
  \caption{GARCH(1, 3) model of Nordea Bank 15min returns}
  \label{tab:nordea_15min_garch}
\end{table}

@@

For the present case with $h = 15\text{min}$, a reasonally good
sampling frequency is 1/30sec, as confirmed by the QQ-plot
\ref{fig:nordea_15min_epOversig_qq} as well as by the autocorrelation
functions of $r_{t+h, h} / \sigma_{t+h, h}$ and of $(r_{t+h, h} /
\sigma_{t+h, h})^2$ which are shown in
\ref{fig:nordea_15min_quotient_acf} and
\ref{fig:nordea_15min_quotient_squared_acf} respectively.
\begin{figure}[htb!]
  \centering
  \subfigure[ACF of $r_{t+h, h} / \sigma_{t+h, h}$]{
    \includegraphics[scale=0.4, clip=true, trim=95 236 118
    200]{../pics/nordea_15min_quotient_acf.pdf}
    \label{fig:nordea_15min_quotient_acf}
  }
  \subfigure[ACF of $(r_{t+h, h} / \sigma_{t+h, h})^2$]{
    \includegraphics[scale=0.4, clip=true, trim=95 236 118
    200]{../pics/nordea_15min_quotient_squared_acf.pdf}
    \label{fig:nordea_15min_quotient_squared_acf}
  }
  \caption{Nordea 15min $r_t/\sigma_t$ and $(r_t/\sigma_t)^2$ ACF.}
\end{figure}

We see in figure \ref{fig:nordea_15min_quotient_acf} and
\ref{fig:nordea_15min_quotient_squared_acf} there is essentially no
autocorrelation in the series $r_{t+h, h} / \sigma_{t+h, h}$ and
$(r_{t+h, h} / \sigma_{t+h, h})^2$, supporting the assumption that the
series is i.i.d Gaussian distributed. As a result, the problem of
modeling $r_{t+h, h}$ boils down to modeling $\sigma_{t+h,
  h}$. Andersen and Bollerslev et al reported that, for the exchange
rates between Deutch mark, yen and dollar, $\ln \sigma_{t+h, h}$ is
gaussian distributed \cite{Andersen03}. This is, however, not the case
for our series in question. In fact, in our case, $\ln \sigma_{t+h,
  h}$ is right skewed (skewness 0.3342) and leptokurtic (kurtosis
6.1006). See figure \ref{fig:nordea_15min_logvol_pdf} and
\ref{fig:nordea_15min_logvol_qq}
\begin{figure}[htb!]
  \centering
  \subfigure[PDF of $\ln\sigma_{t+h,h}$]{
    \includegraphics[scale=0.32, clip=true, trim=52 173 46
    113]{../pics/nordea_15min_logvol_hist.pdf}
    \label{fig:nordea_15min_logvol_pdf}
  }
  \subfigure[QQ-plot of $\ln\sigma_{t+h,h}$]{
    \includegraphics[scale=0.4, clip=true, trim=95 236 118
    220]{../pics/nordea_15min_logvol_qq.pdf}
    \label{fig:nordea_15min_logvol_qq}
  }
  \caption{Nordea 15min $\ln\sigma_{t+h, h}$ unconditional
    distribution}
\end{figure}

@@

The autocorrelation function of the differenced process $w_t$, as shown in
figure \ref{fig:nordea_15min_w_acf}, clearly points to a seasonal $(0,
0, 1)\times(0, 0, 1)$ model: There are only 4 non-zero
autocorrelations in the plot, located at lags 1, 32, 33, 34,
respectively; furthermore, the two at 32 and 34 are approximately
equal. Thus we can write down the model as
\begin{eqnarray*}
  w_t = (1 - \theta B)(1 - \Theta B^s) y_t
\end{eqnarray*}
where $\theta$ and $\Theta$ are parameters to be determined and $y_t$
is a noise process with constant variance $\sigma_y^2$ and mean
0. $y_t$ is often refered to as the residuals.
\begin{figure}[htb!]
  \centering
  \subfigure[ACF of $\ln\sigma_{t+h,h}$]{
    \includegraphics[scale=0.4, clip=true, trim=95 230 112
    235]{../pics/nordea_15min_logvol_acf.pdf}
    \label{fig:nordea_15min_logvol_acf}
  }
  \subfigure[ACF of $w_t$]{
    \includegraphics[scale=0.4, clip=true, trim=95 230 112
    235]{../pics/nordea_15min_w_acf.pdf}
    \label{fig:nordea_15min_w_acf}
  }
  \caption{Nordea 15min $\ln\sigma_{t+h, h}$ and $w_t$ autocorrelations}
\end{figure}

The above $(0, 0, 1)\times(0, 0, 1)$ model has the following
autocovariance structure \cite{BoxJenkins94}:
\begin{eqnarray*}
  \gamma_0 &=& \sigma_y^2 (1 + \theta^2)(1 + \Theta^2) \\
  \gamma_1 &=& -\sigma_y^2\theta(1 + \Theta^2) \\
  \gamma_s &=& -\sigma_y^2\Theta(1 + \theta^2) \\
  \gamma_{s+1} &=& \gamma_{s-1}\;=\;\sigma_y^2\theta\Theta
\end{eqnarray*}
These equations together with the measured autocorrelations make
possible an initial estimate of the parameters $\theta$ and $\Theta$:
\begin{eqnarray*}
  {\rho_{s+1}/\rho_s} &=& {\gamma_{s+1}/\gamma_s} \;=\; -{\theta \over
    1 + \theta^2} \\
  {\rho_{s+1}/\rho_1} &=& {\gamma_{s+1}/\gamma_1} \;=\; -{\Theta \over
    1 + \Theta^2} \\
\end{eqnarray*}
Substituting in the measured values shown in table
\ref{tab:nordea_15min_w_acf},
\begin{table}[htb!]
  \centering
  \begin{tabular}{c|c|c|c}
    $\rho_1$ & $\rho_{s-1}$ & $\rho_s$ & $\rho_{s+1}$ \\
    \hline
    -0.4703 &  0.2053 & -0.4564 &  0.2212
  \end{tabular}
  \caption{Nordea 15min $w_t$ autocorrelations}
  \label{tab:nordea_15min_w_acf}
\end{table}
we get
\begin{eqnarray*}
  \theta &=& 0.6890 \\
  \Theta &=& 0.6378
\end{eqnarray*}
Among the two roots of each of the 2nd order equations in the above,
we have chosen the one in the range $(-1, 1)$ so as to ensure
invertibility of the model \cite{BoxJenkins94}.

@@

If $y_t$ is indeed i.i.d random variables as we assumed, we only need
to work out its unconditional distribution. This is shown in figure
\ref{fig:nordea_15min_y_qq} and \ref{fig:nordea_15min_y_acf}.
\begin{figure}[htb!]
  \centering
  \subfigure[QQ-plot of $y_t$]{
    \includegraphics[scale=0.4, clip=true, trim=95 230 112
    210]{../pics/nordea_15min_y_qq.pdf}
    \label{fig:nordea_15min_y_qq}
  }
  \subfigure[ACF of $y_t$]{
    \includegraphics[scale=0.4, clip=true, trim=95 230 112
    210]{../pics/nordea_15min_y_acf.pdf}
    \label{fig:nordea_15min_y_acf}
  }
  \caption{Nordea 15min $y_t$ QQ-plot and autocorrelations}
\end{figure}
In addition, the measured moments of $y_t$ are shown in table
\ref{tab:nordea_15min_y_moments}:
\begin{table}[htb!]
  \centering
  \begin{tabular}{c|c|c|c}
    mean & variance & skewness & kurtosis \\
    \hline
    0.0012 & 0.0935 & 0.2988 & 6.8691
  \end{tabular}
  \caption{Nordea 15min $y_t$ moments}
  \label{tab:nordea_15min_y_moments}
\end{table}

@@

It turns out that $y_t$ can be well described by a Johnson Su
distribution \cite{Shang2004}:
\[
  y_t = \xi + \lambda\sinh{z_t - \gamma \over \delta}
\]
where $\gamma, \delta, \lambda, \xi$ are parameters to be determined
and $z_t \sim N(0, 1)$. The goodness of fitting is demonstrated in
figure \ref{fig:nordea_15min_y_js_fit} by the
empirical cummulative distribution function in comparison to the
theoretical one.
\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.5, clip=true, trim=112 272 107
    140]{../pics/nordea_15min_y_js_fit.pdf}
    \caption{Nordea 15min $y_t$ fitted to a Johnson Su distribution}
    \label{fig:nordea_15min_y_js_fit}
\end{figure}

>>>>

\documentclass{book}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{wrapfig}
\usepackage{amssymb}
\usepackage{amsfonts, amsmath}
\usepackage{mathrsfs}
\usepackage{enumerate}
\usepackage[bookmarks=true]{hyperref}
\usepackage{bookmark}

@@

\chapter{Introduction}
The mathematics of the financial market has always been a topic that
arouses interest and imagination, and with little doubt, has been
studied from many aspects and in many different ways.

Central to all these studies are the concepts of probability
distributions and correlations. The values of stocks, futures,
options, etc. are stochastic in nature and are governed by the laws of
stochastic processes, which are expressed in terms of probability
distributions and correlations.

On the other hand, the observables of the market are prices, volumes,
turn-over (the amount of money paid in a trade), names of the brokers,
and time of the trades. These quantities don't make much sense
by themselves but do reveal the probabilistic dynamics of the market
when put together and turned into statistics.

Here by dynamics we mean how the value of an asset is affected by its
own history as well as by the histories and current values of other
assets in the market. The influence from the asset's own history is
termed autocorrelations, i.e. correlations in time, while
the influence from other assets are termed cross correlations.

Mathematical models that describe the aforementioned observables in
terms of probability distributions, autocorrelations and cross
correlations give predictive power and help determine a fair price of
a given asset. Hence they are a central topic of the mathematics of
the financial market.

Once a model has been tentatively constructed, it must be
verified. The verification again relies on the statistics of the
observables. Over the years, a number of phenomena have been
consistently reported and become generally accepted. These are referred
to as stylized facts, which are detailed in \S
\ref{sec:StylizedFacts} and constitute a standardized test suit for
model verification.

In \S \ref{sec:FundamentalConcepts} we present a few very
important concepts and notations that we will often refer to in later
chapters. Then the rest of this thesis is organized as follows:

\section{Fundamental Concepts \& Notations}\label{sec:FundamentalConcepts}
This section is a list of a few concepts that may be unfamiliar to the
reader and that we will often refer to later in the thesis:
\begin{itemize}
\item Return. Given a fixed time interval $[t, t']$, for example, a
  day, a week, a month, etc, and a particular asset, for example, a
  share in company ABC, the return of this asset over the time
  interval is defined as
  \begin{eqnarray*}
    r_{t', \Delta t} &=& \ln p_{t'} - \ln p_t \\
    &=& \ln \left(1 + {p_{t'} - p_t \over p_t}\right) \\
    &\approx& {p_{t'} - p_t \over p_t}
  \end{eqnarray*}
  where $p_t$ is the price of the asset at time $t$ and $\Delta t = t'
  - t$ is sometimes called the time-lag of the return. Quite often,
  where confusion is not possible, we will just write $r_t$ to mean
  $r_{t, \Delta t}$, the time lag either does not matter or is clear
  from the context.

\item Autocorrelation. By autocorrelation, denoted $\rho_k$ here, we
  mean the correlation between two temporally separated observations
  of the same time series:
  \begin{eqnarray*}
    \rho_k &=& {
      E\left[(a_t -E(a_t))(a_{t-k} - E(a_{t-k})\right]
      \over
      \sqrt{\text{var}(a_t)}\sqrt{\text{var}(a_{t-k})}
    }
  \end{eqnarray*}
  Here $E(x)$ stands for the expectation value of $x$, and
  $\text{var}(x)$ stands for the variance of $x$. $k$ is called the
  time-lag and is the temporal seperation of the two observations
  measured by the number of observations in between. For example, the
  time-lag between the 1st and 3rd observation is 2.

\item Cross-Correlation Matrix. Correlations between the returns of a
  group of assets are described by the cross-correlation matrix. When
  the asset returns are described by a stable distribution law with
  L\'evy index $\alpha$, the empirical cross-correlation matrix is
  constructed as
  \begin{equation}
    \label{eq:cross-correlation}
    \begin{aligned}
      C_{ij} &= {1 \over T^{2/\alpha}} \sum_{t=1}^T [r_{i,t}-E(r_i)]
      [r_{j,t}-E(r_j)] \\
      &= {1 \over T^{2/\alpha}} RR'
    \end{aligned}
  \end{equation}
  where $r_{i,t}$ is the return of the i-th asset at time t and is
  placed at the entry (i, t) of matrix R; $R'$ denotes the
  transpose of R. In most practical situations, one has abundant data
  for each and every asset. Thus $T \geq N$ is assumed throughout this
  thesis.

\item Auto-regressive processes. A time series $r_t$ is called an
  auto-regressive process of order $p$ and denoted AR(p), if it can be
  written in the following form:
  \begin{eqnarray*}
    r_t &=& \sum_{i=1}^p \phi_i r_{t-i} + a_t
  \end{eqnarray*}
  where, for all $i$, $\phi_i \in (-1, 1)$, and the $a_t$'s are
  independent and identically distributed (iid.) random variables with
  zero mean. Obviously this implies the mean of $r_t$ is zero
  too. Apart from this, their distribution of $a_t$ is not restricted
  to any particular form.

  Of particular interest to this thesis is the AR(1) process:
  \begin{eqnarray*}
    r_t = \phi r_{t-1} + a_t
  \end{eqnarray*}
  Its autocorrelation function $\rho_k = \text{corr}(r_t, r_{t-k})$
  ($k = 0, 1, 2, \cdots$) can be easily shown to fall off
  exponentially:
  \begin{eqnarray*}
    \rho_k &=& {E(r_tr_{t-k}) - E(r_{t-k})E(r_{t}) \over
      \sqrt{\var(r_t) \var(r_{t-k})}} \\
    &=& {\phi E(r_{t-1}r_{t-k}) + E(a_t r_{t-k})
      \over
      \sqrt{\var(r_t) \var(r_{t-k})}
    } \\
    &=& \phi \rho_{k-1}
  \end{eqnarray*}
  where $E(a_t r_{t-k}) = 0$ follows from the fact that any return
  $r_{t1}$ must not depend on disturbances $a_{t2}$ that occur later
  in time. The last equation means $\rho_k$ is a geometric series. Since
  $\rho_0 = 1$, we have
  \begin{eqnarray*}
    \rho_k &=& \phi^k \\
    |\rho_k| &=& e^{k\ln|\phi|} \\
  \end{eqnarray*}
  Note that $|\phi| < 1$ and hence $\ln|\phi| < 0$.

  Although $k$ can only take integer values, it is still useful to
  define a correlation time $\tau$ such that $\phi^\tau = 1/2$. Such a
  quantity is more intuitive and constitutes a measure of
  autocorrelations that is universal and comparable among different
  time series' models. From the definition of $\tau$ we get
  \begin{equation}
    \label{eq:tau_def}
    \begin{aligned}
      \phi^\tau &= 1/2 \\
      \tau &= -{\ln 2 \over \ln\phi} \\
      \phi &= 2^{-1/\tau}
    \end{aligned}
  \end{equation}
  Figure \ref{fig:AR1-autocorrelation} shows the autocorrelation
  function of the AR(1) model. As proven above, this function decays
  exponentially.
  \begin{figure}[htb!]
    \vspace{-15mm}
    \centering
    \includegraphics[scale=0.5, clip=true, trim=113 229 115
    139]{../pics/AR1-autocorrelation.pdf}
    \caption{\footnotesize Autocorrelation function of the AR(1) model
      with $\phi=1/\sqrt{2}$. Red circles: autocorrelations at $k=0, 1,
      2, \cdots$. Blue line: $e^{t\ln\phi}$.}
    \label{fig:AR1-autocorrelation}
  \end{figure}
  For more details about conventional time series' models, see
  \cite{BoxJenkins94}.
\end{itemize}

For purposes of later reference, we also list some notations that may
cause confusion to the reader:
\begin{itemize}
\item $E(x)$ or $\mean{x}$: The expectation value of $x$.
\item $\var(x)$: the variance of $x$.
\item $\text{cov}(x,y)$: the covariance of $x$ and $y$.
\item $\text{corr}(x,y)$: the correlation between $x$ and $y$, i.e.
  \begin{equation*}
    \text{corr}(x,y) = {\text{cov}(x,y) \over \sqrt{\var(x)\var(y)}}
  \end{equation*}
\end{itemize}

\section{Stylized facts}\label{sec:StylizedFacts}
This section presents and explains a few ``stylized facts'',
i.e. phenomena that are widely observed and accepted as true.
\begin{itemize}
\item Fat tails. It has been consistently reported by various studies
  - for example \cite{Potters2003} and \cite{Mantegna2000} - that the
  probability density function (PDF) of stock/index returns are not
  Gaussian. Unlike the Gaussian PDF, which is symmetric and falls off
  very quickly as its argument moves from the center to the outskirts
  (tails), the PDF of stock/index returns are higher on the tails,
  i.e. the probability of large fluctuations is higher than is dictated
  by a Gaussian distribution - in fact, even higher than dictated by
  an exponential function. Empirical studies suggest the PDF of the
  returns follows a power-law on the tails; the exponent of the power
  is around 4, depending on the specific stock/index.

  Figure \ref{fig:FatTail} illustrates this feature.
  \begin{figure}[htb!]
    \centering
    \includegraphics[scale=0.5, clip=true, trim=92 229 110
    140]{../pics/FatTail.pdf}
    \caption{\footnotesize Fat tails. Blue: empirical PDF of 15min returns of
      {\it Nordea Bank}. The returns are computed using paid
      prices on the Stockholm OMX market during the period between
      2013-10-10 and 2014-01-29. Red: Gaussian distribution with equal
      mean and variance.}
    \label{fig:FatTail}
  \end{figure}

\item Non-zero skewness. Apart from fat tails, the PDF of stock/index
  returns are often also skewed. If the skewness is positive
  (negative), the probability of very large positive (negative)
  returns is higher than that of very large negative (positive)
  returns, even though the mean of the returns is 0 or extremely
  close to 0.

  Table \ref{tab:EmpiricalSkewness} lists the skewness of a few
  Swedish stocks traded on the Stockholm OMX market.
  \begin{table}[htb!]
    \footnotesize
    \centering
    \begin{tabular}{c|c|c|c|c|c|c}
      Nordea Bank & Volvo B & Boliden & ABB Ltd & H\&M & Scania
      B & Ericsson B \\
      \hline
      0.1362 & 0.0471 & 0.0567 & -0.0364 & -0.0168 &
      -1.1554 & 0.2086 \\
    \end{tabular}
    \caption{\footnotesize Skewness of Stock Returns. All the returns
      have time-lag of 15 minutes and are computed using paid prices
      between 2013-10-10 and 2014-01-29. }
    \label{tab:EmpiricalSkewness}
  \end{table}

\item Higher-order autocorrelation. 
\end{itemize}

\section{GARCH models}
The autocorrelation function of $\epsilon_t^2$ is given by
(c.f. \cite{Bollerslev87})
$$
\rho_n = \sum_{i=1}^{p \vee q} (\alpha_i + \beta_i) \rho_{n-i}
\;n > p
$$
where $\alpha_i$ with $i > q$ and $\beta_i$ with $i > p$ are taken as
zeros. $p \vee q$ denotes the maximum of p and q. From these
equations, it is clear that the partial autocorrelation function cuts
off at $\max(p, q)$.

\chapter{Price Models}
\section{Nordea Bank}
\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.5, clip=true, trim=103 236 118
  200]{../pics/nordea_price_20120116-20120420.pdf}
  \caption{\footnotesize Nordea Bank paid prices 2012/01/16-2012/04/20}
  \label{fig:Nordea}
\end{figure}
Figure \ref{fig:Nordea} shows the paid prices of Nordea Bank from
2012/01/16 to 2012/04/20.

\subsection{15min Returns}
\subsubsection{GARCH model}
The autocorrelation function (ACF) of $\epsilon_t^2$ are shown in figure
\ref{fig:nordea_15min_acf}. At the absence of GARCH effects, the ACF
will have an asymptotic Gaussian distribution with mean 0 and variance
1/T \cite{Bollerslev86, Bollerslev87}. The first 5 autocorrelations
have considerable sizes and do not fall off as in the Gaussian
case. This observation suggests a GARCH model. Moreover, from
\ref{fig:nordea_15min_vlt_acf}, one sees apparent autocorrelations
between volatilities of consecutive time intervals. This again points
to a GARCH model.
\begin{figure}[htb!]
  \centering
  \subfigure[ACF of squared returns]{
    \includegraphics[scale=0.4, clip=true, trim=95 236 118
    200]{../pics/nordea_15min_acf.pdf}
    \label{fig:nordea_15min_acf}
  }
  \subfigure[ACF of volatilities]{
    \includegraphics[scale=0.4, clip=true, trim=95 236 118
    200]{../pics/nordea_15min_vlt_acf.pdf}
    \label{fig:nordea_15min_vlt_acf}
  }
  \caption{\footnotesize Nordea Bank 15min autocorrelations}
\end{figure}

To obtain the functional form of the conditional distribution of
$\epsilon_t$ \footnote{We use the realized volatility as an estimate of the
conditional volatility. See \cite{Andersen03}.}, we look at the
qq-plot of $\epsilon_t/\sigma_t$ (figure
\ref{fig:nordea_15min_epOversig_qq}), which is consistent with a
standard Gaussian distribution except for a few outliers at the two
ends. Hence the fat tails of the probability density function can be
entirely accounted for by GARCH effects, implying a Gaussian
conditional distribution is appropriate for $\epsilon_t$.
% \begin{figure}[htb!]
\begin{wrapfigure}{r}{0.5\textwidth}
  \centering
    \includegraphics[scale=0.35, clip=true, trim=100 236 118
    200]{../pics/nordea_15min_epOversig_qq.pdf}
  \caption{\scriptsize{QQ-plot of $\epsilon_t / \sigma_t$. $\epsilon_t$ are
    derived from Nordea Bank 15min returns while $\sigma_t$ are realized
    volatilities calculated using 30s returns within each 15min
    interval.}}
  \label{fig:nordea_15min_epOversig_qq}
\end{wrapfigure}

To determine the orders p and q of the GARCH model, we again check
the partial autocorrelation function of $\epsilon_t^2$. As just
mentioned, the first 5 partial autocorrelations are significant. Thus
we can tentatively estimate a GARCH(1, 5) model. Then by applying the
Lagrange multiplier test and accordingly removing insignificant
parameters, we obtain a GARCH(1, 3) model
\ref{tab:nordea_15min_garch}:
\begin{table}[htb!]
  \centering
  \begin{tabular}{c|c|c|c}
    parameter & value & std. error & t statistic \\
    \hline
     $\alpha_0$ &   1.05668e-06 &   3.56871e-07 &    2.96096\\
     \hline
     $\beta_1$  &     0.687212  &   0.0299314   &    22.9596\\
     \hline
     $\alpha_1$ &      0.128614 &   0.0268144   &    4.79643\\
     \hline
     $\alpha_3$ &    0.0568022  &   0.0301503   &    1.88397
  \end{tabular}
  \caption{\footnotesize GARCH(1, 3) model of Nordea Bank 15min returns}
  \label{tab:nordea_15min_garch}
\end{table}

@@

For the present case with $h = 15\text{min}$, a reasonally good
sampling frequency is 1/30sec, as confirmed by the QQ-plot
\ref{fig:nordea_15min_epOversig_qq} as well as by the autocorrelation
functions of $r_{t+h, h} / \sigma_{t+h, h}$ and of $(r_{t+h, h} /
\sigma_{t+h, h})^2$ which are shown in
\ref{fig:nordea_15min_quotient_acf} and
\ref{fig:nordea_15min_quotient_squared_acf} respectively.
\begin{figure}[htb!]
  \centering
  \subfigure[ACF of $r_{t+h, h} / \sigma_{t+h, h}$]{
    \includegraphics[scale=0.4, clip=true, trim=95 236 118
    200]{../pics/nordea_15min_quotient_acf.pdf}
    \label{fig:nordea_15min_quotient_acf}
  }
  \subfigure[ACF of $(r_{t+h, h} / \sigma_{t+h, h})^2$]{
    \includegraphics[scale=0.4, clip=true, trim=95 236 118
    200]{../pics/nordea_15min_quotient_squared_acf.pdf}
    \label{fig:nordea_15min_quotient_squared_acf}
  }
  \caption{\footnotesize Nordea 15min $r_t/\sigma_t$ and $(r_t/\sigma_t)^2$ ACF.}
\end{figure}

We see in figure \ref{fig:nordea_15min_quotient_acf} and
\ref{fig:nordea_15min_quotient_squared_acf} there is essentially no
autocorrelation in the series $r_{t+h, h} / \sigma_{t+h, h}$ and
$(r_{t+h, h} / \sigma_{t+h, h})^2$, supporting the assumption that the
series is i.i.d Gaussian distributed. As a result, the problem of
modeling $r_{t+h, h}$ boils down to modeling $\sigma_{t+h,
  h}$. Andersen and Bollerslev et al reported that, for the exchange
rates between Deutch mark, yen and dollar, $\ln \sigma_{t+h, h}$ is
gaussian distributed \cite{Andersen03}. This is, however, not the case
for our series in question. In fact, in our case, $\ln \sigma_{t+h,
  h}$ is right skewed (skewness 0.3342) and leptokurtic (kurtosis
6.1006). See figure \ref{fig:nordea_15min_logvol_pdf} and
\ref{fig:nordea_15min_logvol_qq}
\begin{figure}[htb!]
  \centering
  \subfigure[PDF of $\ln\sigma_{t+h,h}$]{
    \includegraphics[scale=0.32, clip=true, trim=52 173 46
    113]{../pics/nordea_15min_logvol_hist.pdf}
    \label{fig:nordea_15min_logvol_pdf}
  }
  \subfigure[QQ-plot of $\ln\sigma_{t+h,h}$]{
    \includegraphics[scale=0.4, clip=true, trim=95 236 118
    220]{../pics/nordea_15min_logvol_qq.pdf}
    \label{fig:nordea_15min_logvol_qq}
  }
  \caption{\footnotesize Nordea 15min $\ln\sigma_{t+h, h}$ unconditional
    distribution}
\end{figure}

@@

The autocorrelation function of the differenced process $w_t$, as shown in
figure \ref{fig:nordea_15min_w_acf}, clearly points to a seasonal $(0,
0, 1)\times(0, 0, 1)$ model: There are only 4 non-zero
autocorrelations in the plot, located at lags 1, 32, 33, 34,
respectively; furthermore, the two at 32 and 34 are approximately
equal. Thus we can write down the model as
\begin{eqnarray*}
  w_t = (1 - \theta B)(1 - \Theta B^s) y_t
\end{eqnarray*}
where $\theta$ and $\Theta$ are parameters to be determined and $y_t$
is a noise process with constant variance $\sigma_y^2$ and mean
0. $y_t$ is often refered to as the residuals.
\begin{figure}[htb!]
  \centering
  \subfigure[ACF of $\ln\sigma_{t+h,h}$]{
    \includegraphics[scale=0.4, clip=true, trim=95 230 112
    235]{../pics/nordea_15min_logvol_acf.pdf}
    \label{fig:nordea_15min_logvol_acf}
  }
  \subfigure[ACF of $w_t$]{
    \includegraphics[scale=0.4, clip=true, trim=95 230 112
    235]{../pics/nordea_15min_w_acf.pdf}
    \label{fig:nordea_15min_w_acf}
  }
  \caption{\footnotesize Nordea 15min $\ln\sigma_{t+h, h}$ and $w_t$ autocorrelations}
\end{figure}

The above $(0, 0, 1)\times(0, 0, 1)$ model has the following
autocovariance structure \cite{BoxJenkins94}:
\begin{eqnarray*}
  \gamma_0 &=& \sigma_y^2 (1 + \theta^2)(1 + \Theta^2) \\
  \gamma_1 &=& -\sigma_y^2\theta(1 + \Theta^2) \\
  \gamma_s &=& -\sigma_y^2\Theta(1 + \theta^2) \\
  \gamma_{s+1} &=& \gamma_{s-1}\;=\;\sigma_y^2\theta\Theta
\end{eqnarray*}
These equations together with the measured autocorrelations make
possible an initial estimate of the parameters $\theta$ and $\Theta$:
\begin{eqnarray*}
  {\rho_{s+1}/\rho_s} &=& {\gamma_{s+1}/\gamma_s} \;=\; -{\theta \over
    1 + \theta^2} \\
  {\rho_{s+1}/\rho_1} &=& {\gamma_{s+1}/\gamma_1} \;=\; -{\Theta \over
    1 + \Theta^2} \\
\end{eqnarray*}
Substituting in the measured values shown in table
\ref{tab:nordea_15min_w_acf},
\begin{table}[htb!]
  \centering
  \begin{tabular}{c|c|c|c}
    $\rho_1$ & $\rho_{s-1}$ & $\rho_s$ & $\rho_{s+1}$ \\
    \hline
    -0.4703 &  0.2053 & -0.4564 &  0.2212
  \end{tabular}
  \caption{\footnotesize Nordea 15min $w_t$ autocorrelations}
  \label{tab:nordea_15min_w_acf}
\end{table}
we get
\begin{eqnarray*}
  \theta &=& 0.6890 \\
  \Theta &=& 0.6378
\end{eqnarray*}
Among the two roots of each of the 2nd order equations in the above,
we have chosen the one in the range $(-1, 1)$ so as to ensure
invertibility of the model \cite{BoxJenkins94}.

@@

If $y_t$ is indeed i.i.d random variables as we assumed, we only need
to work out its unconditional distribution. This is shown in figure
\ref{fig:nordea_15min_y_qq} and \ref{fig:nordea_15min_y_acf}.
\begin{figure}[htb!]
  \centering
  \subfigure[QQ-plot of $y_t$]{
    \includegraphics[scale=0.4, clip=true, trim=95 230 112
    210]{../pics/nordea_15min_y_qq.pdf}
    \label{fig:nordea_15min_y_qq}
  }
  \subfigure[ACF of $y_t$]{
    \includegraphics[scale=0.4, clip=true, trim=95 230 112
    210]{../pics/nordea_15min_y_acf.pdf}
    \label{fig:nordea_15min_y_acf}
  }
  \caption{\footnotesize Nordea 15min $y_t$ QQ-plot and autocorrelations}
\end{figure}
In addition, the measured moments of $y_t$ are shown in table
\ref{tab:nordea_15min_y_moments}:
\begin{table}[htb!]
  \centering
  \begin{tabular}{c|c|c|c}
    mean & variance & skewness & kurtosis \\
    \hline
    0.0012 & 0.0935 & 0.2988 & 6.8691
  \end{tabular}
  \caption{\footnotesize Nordea 15min $y_t$ moments}
  \label{tab:nordea_15min_y_moments}
\end{table}

@@

It turns out that $y_t$ can be well described by a Johnson Su
distribution \cite{Shang2004}:
\[
  y_t = \xi + \lambda\sinh{z_t - \gamma \over \delta}
\]
where $\gamma, \delta, \lambda, \xi$ are parameters to be determined
and $z_t \sim N(0, 1)$. The goodness of fitting is demonstrated in
figure \ref{fig:nordea_15min_y_js_fit} by the
empirical cummulative distribution function in comparison to the
theoretical one.
\begin{figure}[htb!]
  \vspace{-18mm}
  \centering
    \includegraphics[scale=0.35, clip=true, trim=112 272 107
    140]{../pics/nordea_15min_y_js_fit.pdf}
    \caption{\footnotesize Nordea 15min $y_t$ fitted to a Johnson Su distribution}
    \label{fig:nordea_15min_y_js_fit}
\end{figure}
